A Shell Model for Real-Time Medical Simulation

### 2010-03-02 MICCAI10/paper.tex
miccai paper v0.0
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@79 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\title{A Shell Model for Real-Time Simulation of Intra-ocular Implant Deployment}
\author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
\institute{INRIA, Alcove team, Lille, France}

@@

\begin{abstract}
With 30 million interventions a year worldwide, cataract surgery is one of the most frequently performed procedures. Yet, no tool currently allows teaching all steps of the procedure without putting patients at risk. A particularly challenging stage of this surgery deals with the injection and deployment of the intra-ocular lens implant. In this paper we propose to rely on shell theory to accurately describe the complex deformations of the implant. Our approach extends the co-rotational method used in finite element analysis of in-plane deformations to incorporate a bending energy. This results in a relatively simple and computationally efficient approach which was applied to the simulation of the lens deployment. This simulation also accounts for the complex contacts that take place during the injection phase. 
%Particular care was given to the validation of the proposed model, its computational efficiency and stability. 
\end{abstract}

\section{Introduction}
According to the latest assessment of the World Health Organization, age related cataract is responsible for $48\,\%$ of world blindness, which represents about 18 million people. Cataract is the clouding of the lens of the eye which impedes the passage of light. Its treatment is a surgical operation, where the opaque lens is removed and replaced by an artificial intra-ocular lens. As people live longer, more surgical interventions are needed. However, current methods of teaching are not compatible with increasing demands of training for this procedure, and also expose patients to complications inherent to the inexperience of the operator. A cataract surgery simulator can offer an efficient alternative teaching method for ophthalmologists, as illustrated in recent reports such as \cite{Feudner09}. The work presented in this paper focuses on the simulation of a crucial step of cataract surgery: the injection and deployment of an artificial lens. To reach this objective, we propose an efficient yet simple formulation for triangular shells, as well as a physically accurate model of the complex (self-)contacts that occur during the injection and deployment of the artificial lens.

@@

Although various shapes of intra-ocular lens implants exist they all share common properties: they are three-dimensional structures, usually made of acrylic
%\footnote{acrylic is the common name for polymethylmethacrylate} 
material, and are much thinner in one direction ($0.13$\,mm) than in the two other ones ($6$ and $13$\,mm). This last characteristic is what classifies these lenses as thin objects, i.e. three-dimensional objects that can be abstracted as two-dimensional structures. Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations. Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects. Most of them fall into one of the following two categories: plate theory or shell theory. Those theories have been a subject of interest in the mechanical community for decades. The difference between these two kinds of structures is very well explained by Liu \textit{et al.}~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations. 
For instance, if we consider the horizontal board of a bookshelf, that board can be approximated as a plate structure and the transversal loads are the weight of the books. A typical deformation of the board is illustrated in figure~\ref{fig-boards} left. Conversely, a shell structure can carry loads in all directions, and therefore can undergo bending, twisting and in-plane deformation (see Figure~\ref{fig-boards} right). 
Considering the folded shape of the implant within the injection device (see Figure \ref{fig-surgery}), a thin plate model would not correctly capture the deformation and therefore a shell formulation was retained to model the intra-ocular lens behaviour.

@@

\section{Simulation of the intra-ocular lens injection}

%Cataract surgery is the removal of the natural lens of the eye (also called \emph{crystalline lens}) that has developed an opacification, which is referred to as a cataract. Metabolic changes of the crystalline lens fibers over the time lead to the development of the cataract and loss of transparency, causing impairment or loss of vision. During cataract surgery, a patient's cloudy natural lens is removed and replaced with a synthetic lens to restore the lens' transparency (figure \ref{fig-surgery}). The last step in the simulation of this surgery is the injection of the implant and its deployment within the lens capsule.

Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens. Prior to starting the surgery, a viscoelastic fluid is introduced into the capsule to facilitate capsulorhexis creation and provide protection during phacoemulsification. This fluid remains in the capsule for the duration of the surgery, including the injection of the implant. Capsulorhexis is the technique used to remove a part of the anterior lens capsule. Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the  fragments of the cortical material. After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about 2 mm) using a foldable intra-ocular lens (see figure \ref{fig-surgery}). The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification. In some cases the implant can be flipped or  the hooks (also called \emph{haptics}) can break. Therefore the simulation of the insertion and deployment of the implant is crucial for achieving a successful surgery.

\begin{figure}[h]
\begin{center}
\includegraphics[height=3.2cm]{images/suction}
\hfill
\includegraphics[height=3.2cm]{images/implant_injection_step_1}
\hfill
\includegraphics[height=3.2cm]{images/implant_injection_step_2}
\caption [Cataract surgery] {Left: removal of the opacified lens by phacoemulsification. Center: insertion of the lens implant which is folded inside the injection device and then deploys within the lens capsule. Right: the lens in place in the capsule.}
\label{fig-surgery}
\end{center}
\end{figure}

To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface. Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (figure \ref{fig-mesh}). This was done by noting the constraints applied by the surgeon to the haptics while inserting the implant within the injection device. During this stage, the haptics are folded onto the implant body, leading to high stresses at the junctions. The lens mesh contains $743$ triangles and $473$ nodes. Models of the injection device and the entire eye anatomy were also created. Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented in table \ref{tab-parameters}.

\begin{table}[h!]
	\begin{center}
		\begin{tabular}{|p{3cm}|p{3cm}|p{3cm}|}
		\hline
		 \centering Young's modulus & \centering Poisson's ratio & \centering Mass density \tabularnewline
		\hline
		\centering $1 MPa$ & \centering $0.42$ & \centering $1.2 g/cm^3$ \tabularnewline
		\hline
		\end{tabular}
	\vspace{0.3cm}
	\caption{Physical parameters of the intra-ocular implant (source: Alcon)}
	\label{tab-parameters}
	\end{center}
\end{table}

The first difficulty is to obtain the folded geometry of the lens within the injection device. This step is not important for the training process and does not need to be interactive. Indeed the surgeon does not always have to prepare the implant as some injection devices are readily available with a folded implant already in place. We simulate the folding process by first folding the haptics onto the implant body. The body was then bent while keeping the haptics inside to obtain the shape described in figure \ref{fig-implantFolding}. The whole process was carried out by applying the necessary forces and boundary conditions on the body and haptics of the implant. The folded implant was then placed into the injection device. The simulation of the injection consists in pushing the intra-ocular implant within the injection device into the lens capsule. During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule. To solve the contacts we use the contact warping method proposed Saupin \emph{et al.} \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models. As adhesions between the haptics and the body is often observed in surgery, friction is also taken into account in the contact response process.

\begin{figure}[!h]
\centering
\includegraphics[height=5cm]{images/IOL}
\includegraphics[height=5cm]{images/mesh_implant}
\caption [Lens implant and its mesh] {An actual intra-ocular implant and the triangular mesh used in our simulations. Notice the higher density of elements in areas where large deformations will take place.}
\label{fig-mesh}
\end{figure}

\begin{figure}[!h]
\centering
\includegraphics[width=4cm]{images/implant_folding1}
\includegraphics[width=3.5cm]{images/implant_folding2} \\
\includegraphics[width=3cm]{images/implant_folding3}
\includegraphics[width=5.5cm]{images/implant_folding4}
\caption [Folding of intra-ocular implant] {Top: intermediary steps of the intra-ocular implant folding. Bottom: fully folded implant ready to be placed into the injection device.}
\label{fig-implantFolding}
\end{figure}

Results of our simulation are illustrated in figure \ref{fig-simu-results}. We can notice the progressive deployment of the implant when it exits the injector.  The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule. Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01$\,s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).

\begin{figure}[!h]
\centering
\includegraphics[width=4cm]{images/surgery1}
\hfill
\includegraphics[width=4cm]{images/surgery2}
\hfill
\includegraphics[width=4cm]{images/surgery3} \\
\vspace{0.1cm}
\includegraphics[width=4cm]{images/simu1}
\hfill
\includegraphics[width=4cm]{images/simu2}
\hfill
\includegraphics[width=4cm]{images/simu3}
\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule. Top: images from a real cataract surgery, courtesy of Dr. Tarek Youssef, ophtalmologist in Canada. Below: our simulation of the implant's deployment.}
\label{fig-simu-results}
\end{figure}

@@

\section*{Acknowledgments}
We are very grateful to J\'er\'emy Ringard for his invaluable help in improving the eye anatomical model, to J\'er\'emie Dequidt and Damien Marchal for understanding Blender, to Fr\'ederick Roy for his help with SOFA, and Nadia Boubchir for her persistence in motivating us to obtain these results in time.

\bigskip

We would also like to thank INRIA and the SOFA project, as well as CSIRO Preventative Health National Research Flagship, ICTC, The Australian e-Health Research Centre for their fundings.

>>>>

\title{A Shell Model for Real-Time Medical Simulation}
\author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
\institute{INRIA, Alcove team, Lille, France}

@@

\begin{abstract}
Contributions: Shell model with real-time implementation + method for fitting medical data

\end{abstract}

\section{Introduction}

@@

Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations. Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects. Most of them fall into one of the following two categories: plate theory or shell theory. Those theories have been a subject of interest in the mechanical community for decades. The difference between these two kinds of structures is very well explained by Liu \textit{et al.}~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations. 
For instance, if we consider the horizontal board of a bookshelf, that board can be approximated as a plate structure and the transversal loads are the weight of the books. A typical deformation of the board is illustrated in figure~\ref{fig-boards} left. Conversely, a shell structure can carry loads in all directions, and therefore can undergo bending, twisting and in-plane deformation (see Figure~\ref{fig-boards} right). 
Considering the folded shape of the implant within the injection device (see Figure \ref{fig-surgery}), a thin plate model would not correctly capture the deformation and therefore a shell formulation was retained to model the intra-ocular lens behaviour.

@@

@@
